---
title: "4.7 Exercises"
output:
  html_document:
    toc: True
---

# Exercise 10

This questions should be answered using the **Weekly** data set, which is part of the **ISLR** package. This data is similar in nature to the **Smarket** data from this chapter's lab, except that it contains 1,089 weekly returns for 21 years, from the beginning of 1990 to the end of 2010.

(a) Produce some numerical and graphical summaries of the **Weekly** data. Do there appear to be any patterns?

```{r}
library(ISLR)
fix(Weekly)
```

```{r}
summary(Weekly)
```

*As you would expect, the lag variables and the* **Today** *variable all share the same distributions.*

```{r}
605 / (484 + 605)
```

*There are good deal more* **Up** *days than there are* **Down** *days*. 55.56% *of the day's are recorded as a positive weekly return.*


```{r}
plot(Weekly$Today)
```

```{r}
hist(Weekly$Today)
```

*At a first glance, the weekly returns are pretty normally distributed around* 0, *with no real trends over time.*

```{r}
plot(Weekly$Volume)
```

*The* **Volume** *of shares traded increases in an exponential fashion over time, so there has definitely been consistent market growth.*

(b) Use the full data set to perform a logistic regression with **Direction** as the response and the five lag variables plus **Volume** as predictors. Use the summary function to print the results. Do any of the predictors appear to be statistically significant? If so, which ones?

```{r}
glm.fit = glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Weekly, family = binomial)
summary(glm.fit)
```

*The only statistically significant predictor appears to be* **Lag2**, *with a* p-*value of* 0.0296.

(c) Compute the confusion matrix and overall fraction of correct predictions. Explain what the confusion matrix is telling you about the types of mistakes made by logistic regression.

```{r}
glm.prob = predict(glm.fit, type = 'response')
glm.pred = rep('Down', length(glm.prob))
glm.pred[glm.prob > .5] = 'Up'
table(glm.pred, Weekly$Direction)
```

*At a probability cutoff of* 0.5, *this model is very optimistic. It predicts that* 987 *of the weeks from the training data should have a positive return* (*versus the* 605 *weeks of the training data that actually have a positive return*). 987 *weeks corresponds to* 90.6% *of the data* (*versus the* 55.6% *that actually had a positive return*).

(d) Now fit the logistic regression omdel using a training data preiod from 1990 to 2008, with **Lag2** as the only predictor. Compute the confusion matrix and the overall fraction of correct predictions for the held out data (that is, the data form 2009 and 2010).

```{r}
train = Weekly$Year < 2009
summary(Weekly[train,])
```

```{r}
x = 441 + 544
x / (484 + 605)
544 / x
```

*The training data consists of* 90.4% *of the whole data set, leaving* 9.6% *of the data to use as test data. The percentage of training data with a positive weekly return is close to the overall percentage at* 55.2%.

```{r}
summary(Weekly[!train,])
```

```{r}
61 / (43 + 61)
```

*The percentage of positive weekly returns is a bit higher for the test data, at* 58.7%. *Conversely, the percentage of negative weekly returns is* 41.3%.

```{r}
glm.fit = glm(Direction ~ Lag2, data = Weekly[train,], family = binomial)
summary(glm.fit)
```

```{r}
glm.prob = predict(glm.fit, Weekly[!train,], type = 'response')
glm.pred = rep('Down', length(glm.prob))
glm.pred[glm.prob > .5] = 'Up'
table(glm.pred, Weekly[!train, 'Direction'])
```

*This model is also quite optimistic, predicting* 86.5% *of the test data to have positive weekly returns.*

```{r}
mean(glm.pred == Weekly[!train, 'Direction'])
```

*The model is correct on the test data* 62.5% *of the time, which is only a bit better than a dummy model of predicting only* **Up**. *The dummy model would have a test accuracy rate of* 58.7%.

(e) Repeat (d) Using LDA.

```{r}
library(MASS)
```

```{r}
lda.fit = lda(Direction ~ Lag2, data = Weekly[train,])
lda.fit
```

```{r}
lda.pred  = predict(lda.fit, Weekly[!train,])
lda.class = lda.pred$class
table(lda.class, Weekly[!train, 'Direction'])
```

```{r}
mean(lda.class == Weekly[!train, 'Direction'])
```

*The LDA model performs exactly the same as the logistic regression model.*

(f) Repeat (d) Using QDA.

```{r}
qda.fit = qda(Direction ~ Lag2, data = Weekly[train,])
qda.fit
```

```{r}
qda.pred  = predict(qda.fit, Weekly[!train,])
qda.class = qda.pred$class
table(qda.class, Weekly[!train, 'Direction'])
```

*The QDA model is worse than the logistic regression and LDA models. It predicts all of the training data to be a positive weekly return, which yields a training accuracy rate of* 58.7% - *the same as the dummy model of only predicting* **Up**.

(g) Repeat (d) using KNN with K = 1.

```{r}
library(class)
```

```{r}
train.X = data.frame(Weekly$Lag2[train])
test.X  = data.frame(Weekly$Lag2[!train])
train.Direction = Weekly$Direction[train]
knn.pred = knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, Weekly$Direction[!train])
```

```{r}
mean(knn.pred == Weekly[!train, 'Direction'])
```

*The KNN classifier with* $K = 1$ *performs poorly. It has a relatively high test error rate, and it does not even have better performance on predicting specifically for* **Up** *or* **Down**.

(h) Which of these methods appear to provide the best results on this data?

*Of these methods, the logistic regression and LDA models perform the best on this data.*

(i) Experiment with different combinations of predictors, including possible transformations and interactions for each of the methods. Report the variables, method, and associated confusion matrix that appears to provide the best results on the hold out data. Note that you should also experiment with values for *K* in the KNN classifier.

```{r}
train.X = Weekly[ train, c('Lag1', 'Lag2', 'Lag3', 'Lag4', 'Lag5')]
test.X  = Weekly[!train, c('Lag1', 'Lag2', 'Lag3', 'Lag4', 'Lag5')]
knn.pred = knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, Weekly$Direction[!train])
```

*Using all of the lag predictors with* $K = 1$ *KNN yields almost the exact same result as just using* **Lag2**.

```{r}
knn.pred = knn(train.X, test.X, train.Direction, k = 2)
table(knn.pred, Weekly$Direction[!train])
```

$K = 2$ *is a little bit worse, but still very similar.*

```{r}
for (k in 3:8) {
  knn.pred = knn(train.X, test.X, train.Direction, k = k)
  print(c(k, mean(knn.pred == Weekly[!train, 'Direction'])))
}
```

*KNN with* $K = 4$ *seems to be pretty decent with* 60.6% *test accuracy. However, this performance is still worse than the LDA using* **Lag2**.

```{r}
table(knn(train.X, test.X, train.Direction, k = 4), Weekly[!train, 'Direction'])
```

*Like the other KNN results we have looked at, this model is not as optimistic as the logistic regression, LDA, and QDA. However, it does still predict most of the data to be a positive weekly return.*

```{r}
glm.fit = glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5, data = Weekly[train,], family = binomial)
summary(glm.fit)
```

```{r}
glm.prob = predict(glm.fit, Weekly[!train,], type = 'response')
glm.pred = rep('Down', sum(!train))
glm.pred[glm.prob > .5] = 'Up'
table(glm.pred, Weekly[!train, 'Direction'])
```

```{r}
mean(glm.pred == Weekly[!train, 'Direction'])
```

*The logistic regression using all of the lag predictors doesn't have great performance. It is has a higher test error rate than the dummy model.*

```{r}
glm.fit = glm(Direction ~ Lag2 + Volume, data = Weekly[train,], family = binomial)
glm.prob = predict(glm.fit, Weekly[!train,], type = 'response')
glm.pred = rep('Down', sum(!train))
glm.pred[glm.prob > .5] = 'Up'
mean(glm.pred == Weekly[!train, 'Direction'])
```

```{r}
glm.fit = glm(Direction ~ I(Lag2^2), data = Weekly[train,], family = binomial)
glm.prob = predict(glm.fit, Weekly[!train,], type = 'response')
glm.pred = rep('Down', sum(!train))
glm.pred[glm.prob > .5] = 'Up'
mean(glm.pred == Weekly[!train, 'Direction'])
```

*I can't find anything with a better performance on the test data than the original logistic regression. Note that using transformations like* **log()** *or* **sqrt()** *are not available on the lag predictors since they contain negative values.*

---

# Exercise 11

In this problem, you will develop a model to predict whether a given car gets high or low gas mileage based on the **Auto** data set.

```{r}
fh = 'D:/GoogleDrive/Introduction to Statistical Learning with Applications in R/data-sets/Auto.csv'
Auto = read.csv(file = fh, header = TRUE, na.strings = '?')
```

(a) Create a binary variable, **mpg01**, that contains a 1 if **mpg** contains a value above its median, and a 0 if **mpg** contains a value below its median. You can compute the median using the **median()** function. NOte you may find it helpful to use the **data.frame()** function to create a single data set containing both **mpg01** and the other **Auto** variables.

```{r}
Auto$mpg01 = 0
Auto$mpg01[Auto$mpg > median(Auto$mpg)] = 1
```

(b) Explore the data graphically in order to investigate the association between **mpg01** and the other features. Which of the other features seem most likely to be useful in predicting **mpg01**? Scatterplots and boxplots may be useful tools to answer this question. Describe your findings.

```{r}
par(mfrow=c(3,3))
for (c in names(Auto)[-c(1,10)]) {
  boxplot(Auto$mpg01, Auto[,c], xlab = 'mpg', ylab = c)
}
```

```{r}
idx0 = Auto$mpg01 == 0
idx1 = !idx0
par(mfrow = c(3,3))
xlim = c(min(Auto$mpg), max(Auto$mpg))
for (c in names(Auto)[-c(1,10)]) {
  if (c == 'name') {
    ylim = c(0, length(unique(Auto[,c])))
  } else {
    ylim = c(min(Auto[,c], na.rm = TRUE), max(Auto[,c], na.rm = TRUE))
  }
  plot(Auto$mpg[idx0], Auto[idx0,c], xlab = 'mpg', ylab = c, col = 'red', xlim = xlim, ylim = ylim)
  points(Auto$mpg[idx1], Auto[idx1,c], col = 'blue')
}
```

*It looks like the number of cylinders, the engine displacement, the engine horsepower, and the vehicle weight are all important predictors. The 0 to 60 acceleration and model year of the vehicle may also be useful predictors.*

(c) Split the data into a training set and a test set.

```{r}
set.seed(1)
Auto.less = Auto[Auto$mpg01 == 0,]
Auto.more = Auto[Auto$mpg01 == 1,]
sample.index = function(df, percent) {
  len = dim(df)[1]
  idx = rep(FALSE, len)
  for (i in sample(len, size = floor(percent * len), replace = FALSE)) idx[i] = TRUE
  return(idx)
}
less.test = sample.index(Auto.less, .1)
more.test = sample.index(Auto.more, .1)

Auto.test  = rbind(Auto.less[ less.test,], Auto.more[ more.test,])
Auto.train = rbind(Auto.less[!less.test,], Auto.more[!more.test,])
```

*We'll just take 10% of the data pseudo-randomly. We'll randomly draw 10% of the observations with* **mpg** *greater than it's median and another random draw for 10% of the observations with* **mpg** *less than it's median. That data set will be the test set. The rest of the data will be the training set.*

(d) Perform LDA on the training data in order to predict **mpg01** using the variables that seemed most associated with **mpg01** in (b). What is the test error of the model obtained?

```{r}
form = as.formula(mpg01 ~ cylinders + displacement + horsepower + weight + acceleration + year)
lda.fit = lda(form, data = Auto.train)
lda.fit
```

```{r}
lda.class = predict(lda.fit, Auto.test)$class
table(lda.class, Auto.test$mpg01)
mean(lda.class == Auto.test$mpg01)
```

*The test error rate obtained is* 5.13%. *It seems the selected predictors work well for the LDA method here.*

(e) Repeat (d) using QDA.

```{r}
qda.fit = qda(form, data = Auto.train)
qda.fit
```

```{r}
qda.class = predict(qda.fit, Auto.test)$class
table(qda.class, Auto.test$mpg01)
mean(qda.class == Auto.test$mpg01)
```

*Interestingly, the QDA method has the same test error rate as LDA. However, it tends to prefer to classify* **mpg01** *as* 0, *whereas the LDA result tends to classify* **mpg01** *as* 1.

(f) Repeat (d) using logistic regression.

```{r}
glm.fit = glm(form, data = Auto.train, family = binomial)
summary(glm.fit)
```

```{r}
glm.prob = predict(glm.fit, Auto.test, type = 'response')
glm.pred = rep(0, dim(Auto.test)[1])
glm.pred[glm.prob > .5] = 1
table(glm.pred, Auto.test$mpg01)
mean(glm.pred == Auto.test$mpg01)
```

*The logistic regression method yields, again, the same test error rate. This time though, the incorrect classifications are split between* **mpg01** *of* 0 *and* 1.

(g) Repeat (d) using KNN with multiple values for *K*. Which value of *K* seems to perform the best on this data set?

```{r}
acc.rate = 1:10
x.cols = c('cylinders', 'displacement', 'horsepower', 'weight', 'acceleration', 'year')
train.X = scale(Auto.train[,x.cols])
test.X  = scale(Auto.test[ ,x.cols])

# knn does not allow for any NaN's in the data
idx = complete.cases(train.X)

for (k in 1:10) {
  knn.pred = knn(train.X[idx,], test.X, Auto.train$mpg01[idx], k = k)
  acc.rate[k] = mean(knn.pred == Auto.test$mpg01)
}
df = data.frame(k = 1:10, accuracy = acc.rate, error = 1 - acc.rate)
df
```

*With* $K = 6$, *the KNN method perfectly predicts the test data.*

---

# Exercise 12

This problem involves writing functions.

(a) Write a function, **Power()**, that prints out the result of raising 2 to the 3rd power. In other words, your function should compute $2^3$ and print out the results.
*Hint: Recall that* **x^a** *raises* **x** *to the power* **a**. *Use the* **print()** *function to output the result.*

```{r}
Power = function () {
  print(2^3)
}
Power()
```

(b) Create a new function, **Power2()**, that allows you to pass *any* two numbers, **x** and **a**, and prints out the value of **x^a**. You can do this by beginning your function with the line
```Power2=function(x,a){```
You should be able to call your function by entering, for instance,
```Power2(3,8)```
on the command line. This should output the value of $3^8$, namely, 6,561.

```{r}
Power2 = function (x,a) {
  print(x^a)
}
Power2(3,8)
```

(c) Using the **Power2()** function that you just wrote, compute $10^3$, $8^{17}$, and $131^3$.

```{r}
Power2(10,3)
Power2(8,17)
Power2(131,3)
```

(d) Now create a new function, **Power3()**, that actually *returns* the result **x^a** as an R object, rather than simply printing it to the screen. That is, if you store the value **x^a** in an object called **result** within your function, then you can simply **return()** this result, using the following line:
```return(result)```
The line above should be the last line in your function, before the **}** symbol.

```{r}
Power3 = function (x,a) return(x^a)
result = Power3(2,3)
result
```

(e) Now using the **Power3()** function, create a plot of $f(x) = x^2$. The *x*-axis should display a range of integers from 1 to 10, and the *y*-axis should display $x^2$. Label the axes appropriately, and use an appropriate title for the figure. Consider displaying either the *x*-axis, the *y*-axis, or both on the log-scale. You can do this by using **log='x'**, **log='y'**, or **log='xy'** as arguments to the **plot()** function.

```{r}
x = 1:10
plot(x, Power3(x,2), main = 'x^2 v. x', xlab = 'x', ylab = 'x^2')
```

```{r}
x = 1:10
plot(x, Power3(x,2), main = 'log(x^2) v. x', xlab = 'x', ylab = 'log(x^2)', log = 'y')
```

```{r}
x = 1:10
plot(x, Power3(x,2), main = 'x^2 v. log(x)', xlab = 'log(x)', ylab = 'x^2', log = 'x')
```

```{r}
x = 1:10
plot(x, Power3(x,2), main = 'log(x^2) v. log(x)', xlab = 'log(x)', ylab = 'log(x^2)', log = 'xy')
```

(f) Create a function, **PlotPower()**, that allows you to create a plot of **x** against **x^a** for a fixed **a** and for a range of values **x**. For instance, if you call
```PlotPower(1:10,3)```
then a plot should be created with an *x*-axis taking on values 1,2,...,10, and a *y*-axis taking on values $1^3,2^3,...,10^3$.

```{r}
PlotPower = function (x,a) plot(x, x^a, main = sprintf('x^%i v. x', a), xlab = 'x', ylab = sprintf('x^%i', a))
PlotPower(1:10, 3)
```

---

# Exercise 13

Using the **Boston** data set, fit classification models in order to predict whether a given suburb has a crime rate above or below the median. Explore logistic regression, LDA, and KNN models using various subsets of the predictors. Describe your findings.

```{r}
Boston$crim01 = 0
Boston$crim01[Boston$crim > median(Boston$crim)] = 1
```

*First we make a new variable in the Boston data set that indicates whether the crime rate is above or below the median. We will call this variable* **crim01**. *A value of* 0 *will indicate that the suburb's crime rate is less than or equal to the median. A value of* 1 *will indicate that the suburb's crime rate is above the median.*

```{r}
glm.fit = glm(crim01 ~ ., data = Boston[,-1], family = binomial)
summary(glm.fit)
```

*Next we will explore a logistic regression using all of the predictors, since R will provide statistical significance for each predictor. Of course, we want to exclude* **crim** *from the fit. If you already have access to the crime rate itself, then "predicting" whether or not the crime rate is above the median is trivial.*

*We can see from the fit that the proportion of residential land zoned* (**zn**), *the nitrogen oxides concentration* (**nox**), *the weighted mean of distances to five Boston employment centres* (**dis**), *the index of accessibility to radial highways* (**rad**), *the full-value property-tax rate* (**tax**), *the pupil-teacher ratio* (**ptratio**), *the proportion of blacks by town* (**black**), *and the median value of owner occupied homes* (**medv**) *all have a p-value less than* 5%, *so these are the predictors that we will use for our models.*

```{r}
set.seed(1)
cols = c('crim01', 'zn', 'nox', 'dis', 'rad', 'tax', 'ptratio', 'black', 'medv')
Boston.less = Boston[Boston$crim01 == 0,]
Boston.more = Boston[Boston$crim01 == 1,]

less.test = sample.index(Boston.less, .1)
more.test = sample.index(Boston.more, .1)

Boston.test  = rbind(Boston.less[ less.test, cols], Boston.more[ more.test, cols])
Boston.train = rbind(Boston.less[!less.test, cols], Boston.more[!more.test, cols])
```

*Using the function we defined in Exercise* 11c, *we split the Boston data set into a test set, consisting of* 10% *of the below-median-data and* 10% *of the above-median-data, and a training set, consisting of the rest of the data.*

```{r}
glm.fit = glm(crim01 ~ ., data = Boston.train, family = binomial)
glm.prob = predict(glm.fit, Boston.test, type = 'response')
glm.pred = rep(0, dim(Boston.test)[1])
glm.pred[glm.prob > .5] = 1
table(glm.pred, Boston.test$crim01)
mean(glm.pred == Boston.test$crim01)
```

*The resulting logistic regression has a* 6% *test error rate - not bad.*

```{r}
lda.fit = lda(crim01 ~ ., data = Boston.train)
lda.pred = predict(lda.fit, Boston.test)$class
table(lda.pred, Boston.test$crim01)
mean(lda.pred == Boston.test$crim01)
```

*The linear discriminant analysis also has a* 6% *test error rate. Seems like an amusing coincidence that these two methods yield the same test error rate in both this exercise and in Exercise* 11.

```{r}
qda.fit = qda(crim01 ~ ., data = Boston.train)
qda.pred = predict(qda.fit, Boston.test)$class
table(qda.pred, Boston.test$crim01)
mean(qda.pred == Boston.test$crim01)
```

*Again, a test error rate of* 6%. *And again, amusing. In this case the confusion matrix is the same for both LDA and QDA. It makes you wonder if the same observations were misclassified.*

```{r}
which(lda.pred != Boston.test$crim01)
which(qda.pred != Boston.test$crim01)
```

*LDA and QDA yield almost the exact same results, but not quite.*

```{r}
acc.rate = rep(0,10)
for (k in 1:10) {
  knn.pred = knn(scale(Boston.train[,-1]), scale(Boston.test[,-1]), Boston.train$crim01, k = k)
  acc.rate[k] = mean(knn.pred == Boston.test$crim01)
}
data.frame(k = 1:10, accuracy = acc.rate, error = 1 - acc.rate)
```

*With a value of* 6*or* 9 *for* K, *KNN has a test error rate of* 0%, *perfectly predicting the test data.*
